Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2018-06-02 LaTeX/Turnover_bioRXiv_revision.tex
revision 2 complete by matt

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath}%amssymb

@@

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage[outdir=./]{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}
%% Include all macros below

%for figure alignment
\usepackage[export]{adjustbox}

@@

%reference equations in appendices
\usepackage{xr}
\externaldocument{S1Appendix_revised}
\externaldocument{S2Appendix_revised}
\externaldocument{S3Appendix_revised}

%for figures
%\usepackage{epstopdf}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Haploid selection, sex-ratio bias, and transitions between sex-determining systems} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Michael Francis Scott\textsuperscript{1\Yinyang*},
Matthew Miles Osmond\textsuperscript{2\Yinyang},
Sarah Perin Otto\textsuperscript{2}
\\
\bigskip
\textbf{1} UCL Genetics Institute, Department of Genetics, Evolution and Environment, University College London, Gower Street, London WC1E 6BT 
\\
\textbf{2} Department of Zoology, University of British Columbia, \#4200 - 6270 University Boulevard, Vancouver, BC, Canada V6T 1Z4
\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
%currently 308 words
\section*{Abstract}
Sex determination is remarkably dynamic; many taxa display shifts in the location of sex-determining loci or the evolution of entirely new sex-determining systems. 
Predominant theories for why we observe such transitions generally conclude that novel sex-determining systems are favoured by selection if they equalise the sex ratio or increase linkage with a locus that experiences different selection in males vs.\ females. 
We use population genetic models to extend these theories in two ways: 
(1) We consider the dynamics of loci very tightly linked to the ancestral sex-determining loci, e.g., within the non-recombining region of the ancestral sex chromosomes. 
Variation at such loci can favour the spread of new sex-determining systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or even unlinked) to the locus under selection. %, which is \textcolor{red}{not expected from previous theory}. 
(2) We consider selection upon haploid genotypes either during gametic competition (e.g., pollen competition) or meiosis (i.e., non-Mendelian segregation), which can cause the zygotic sex ratio to become biased. 
Haploid selection can drive transitions between sex-determining systems without requiring selection to act differently in diploid males vs.\ females. 
With haploid selection, we find that transitions between male and female heterogamety can evolve where linkage with the sex-determining locus is either strengthened or weakened. 
%Furthermore, we find that the selective strengths exerted by the sex-ratio and associations with selected alleles are often equivalent. 
Furthermore, we find that sex-ratio biases may increase or decrease with the spread of new sex chromosomes, which implies that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio.
In fact, under many conditions, we find that transitions in sex determination are favoured equally strongly in cases where the sex ratio bias increases or decreases.
%Furthermore, we find that the selective forces to equalize the sex ratio are equally important to selective forces that generate skewed sex ratios when accounting for the spread of new sex chromosomes.
%This allows sex-ratio biases to increase or decrease with the spread of new sex chromosomes and implies that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio. 
Overall, our models predict that sex determination systems should be highly dynamic, particularly when haploid selection is present, consistent with the evolutionary lability of this trait in many taxa.
%Overall, our models reveal that transitions between sex-determining systems, particularly transitions where the heterogametic sex changes, can be driven by loci in previously unpredicted genomic locations that experience selection during diploid and/or haploid phases.
%Our results predict conditions under which sex-determining systems are likely to be labile and draw connections with sex-ratio evolution.

@@

In each generation, we census the genotype frequencies in male and female gametes before gametic competition. 
%Below, it is convenient to follow the ancestral equilibrium frequency of the $A$ allele among female gametes (eggs), $\hat{p}^\female_X$, and among X-bearing, $\hat{p}^\male_X$, and among Y-bearing, $\hat{p}^\male_Y$, male gametes (sperm/pollen). 
%We also track the ancestral fraction of male gametes that are Y-bearing, $\hat{q}_{Y}$, which may deviate from $1/2$ due to meiotic drive in males. 
%
After gametic competition, conjugation between male and female gametes occurs at random.
The resulting zygotes develop as males or females, depending on their genotypes at the $\mathbf{X}$ and $\mathbf{M}$ loci. %(and the $\mathbf{M}$ genotype of their mother in the case of maternal control).
Diploid males and females then experience viability and/or individual-based fertility selection, with relative fitnesses $w_{AA}^{\circ}$, $w_{Aa}^{\circ}$, and $w_{aa}^{\circ}$. %where $g$ is the diploid genotype at the $\mathbf{A}$ locus ($g \in \{AA, Aa, aa\}$).  
We do not consider fertility selection that depends on the mating partner, e.g., sexual selection with variation in choosiness. 
The next generation of gametes is produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination (i.e., an odd number of cross-overs) occurs between loci $\mathbf{X}$ and $\mathbf{A}$ with probability $r$, between loci $\mathbf{A}$ and $\mathbf{M}$ with probability $R$, and between loci $\mathbf{X}$ and $\mathbf{M}$ with probability $\rho$.
Any linear order of the loci can be modelled with appropriate choices of $r$, $R$, and $\rho$ (see Fig \ref{fig:model_outline}A and \nameref{tab:chisubstitutions}). 
Our model is entirely deterministic and hence ignores chance fluctuations in allele frequencies due to genetic drift.
%During the , the $\mathbf{A}$ locus can experience sex-specific gametic competition, diploid selection, and/or meiotic drive.

@@

In the ancestral population, it is convenient to follow the frequency of the $A$ allele among female gametes (eggs), $p^\female_X$, and among X-bearing, $p^\male_X$, or among Y-bearing, $p^\male_Y$, male gametes (sperm/pollen). 
We also track the fraction of male gametes that are Y-bearing, $q$, which may deviate from $1/2$ due to meiotic drive in males. 
We consider only equilibrium frequencies of alleles, $\hat{p}^\circ_i$, and Y-bearing male gametes, $\hat{q}_{Y}$, when determining the invasion of new sex-determining factors.
We use $\zeta$ to measure the sex ratio (fraction male) among zygotes, which is determined by the allele frequencies and haploid selection coefficients (\nameref{tab:meanfitnesses}).

@@

\begin{table}[!ht]
\centering
\smallskip
%\caption{Parameters determining invasion (equation \ref{eq:charpoly_neoY}) for neo-Y or neo-W alleles}
\caption{ {\bf Parameters determining invasion of mutant neo-Y and neo-W alleles into an ancestrally XY system}}
\begin{tabular}{l}
\hline\hline
   \noalign{\vskip 0.5ex}
   $m$ is a neo-Y ($k=0$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\Lambda_{Y'A}^{(XY)} = {\left( 2 \zeta \right)}^{-1} \left[\hat{p}_X^\female w_{A}^{\female} w_{A}^{\male} w_{AA}^{\male} + (1-\hat{p}_X^\female) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} (1+\alpha_{\Delta}^{\male}) \right]/ \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\male}_{D} \right) $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\Lambda_{Y'a}^{(XY)} = {\left( 2 \zeta \right)}^{-1} \left[(1-\hat{p}_X^\female) w_{a}^{\female} w_{a}^{\male} w_{aa}^{\male} + \hat{p}_X^\female w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male}(1 - \alpha_{\Delta}^{\male}) \right]/ \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\male}_{D} \right) $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{Y'A}^{(XY)} = R {\left( 2 \zeta \right)}^{-1} \left[ (1-\hat{p}_X^\female) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} (1+\alpha_{\Delta}^{\male}) \right]/  \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\male}_{D} \right)   $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{Y'a}^{(XY)} = R {\left( 2 \zeta \right)}^{-1} \left[   \hat{p}_X^\female w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male} (1 - \alpha_\Delta^{\male}) \right]/ \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\male}_{D} \right)  $\\ [1ex] \hline 
  \noalign{\vskip 0.5ex}
  $m$ is a neo-W ($k=1$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\Lambda_{W'A}^{(XY)} = {\left[ 2 (1 - \zeta) \right]}^{-1} \left[ \bar{p}^{\male} w_{A}^{\male} w_{A}^{\female} w_{AA}^{\female}+(1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female}(1+\alpha_{\Delta}^{\female})\right]/ \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\female}_{D} \right) $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\Lambda_{W'a}^{(XY)} = {\left[ 2 (1 - \zeta) \right]}^{-1} \left[ (1-\bar{p}^{\male}) w_{a}^{\male} w_{a}^{\female} w_{aa}^{\female}+\bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female}(1-\alpha_{\Delta}^{\female})\right] / \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\female}_{D} \right) $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{W'A}^{(XY)} = R {\left[ 2 (1 - \zeta) \right]}^{-1} \left[ (1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female} (1+\alpha_{\Delta}^{\female}) \right] / \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\female}_{D} \right)  $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{W'a}^{(XY)} = R {\left[ 2 (1 - \zeta) \right]}^{-1} \left[ \bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female} (1-\alpha_{\Delta}^{\female}) \right] / \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\female}_{D} \right) $ \\ [1ex]
  \hline \hline 
   \end{tabular}
   \begin{flushleft} 
$\bar{p}^{\male}=(1-\hat{q}_{Y})\hat{p}_X^\male + \hat{q}_{Y}\hat{p}_Y^\male$ is the average frequency of the $A$ allele among X- and Y-bearing male gametes.
$\hat{q}_{Y}$ is the frequency of Y-bearing male gametes. 
%      \item $R$ is the probability of recombination between loci $\mathbf{A}$ and $\mathbf{M}$.
$\zeta$ is the zygotic sex ratio (fraction male).
$\bar{w}^{\circ}_{D}$ is the mean fitness of diploids of sex $\circ$ $\in\{\female,\male\}$.
$\bar{w}_H^\circ$ is the mean fitness of haploids from sex $\circ$, see \nameref{tab:meanfitnesses}.
$R$ is the rate of recombination between the neo-sex-determiner and the selected locus.
Selection terms ($w_i^\circ$, $\alpha_\Delta^\circ$) are described in Table \ref{tab:fitnesstable}. 
\end{flushleft}
  \label{tab:haplotype_growth}
\end{table}

@@

Neo-W alleles, on the other hand, can invade an ancestral XY system (the full invasion conditions are given in \nameref{app:inv_cond}; equations \ref{Binvasion} and \ref{Ainvasion}).
Invasion occurs when neo-W females can have higher fitness than the XX females in the ancestral population.
Neo-W invasion is possible under all forms of selection that can maintain a polymorphism (sexually-antagonistic selection, overdominance, ploidally-antagonistic selection, or some combination, e.g., \nameref{fig:temporalOverdominance}, \nameref{fig:SexAntagTighterMaleDrive}, and \nameref{fig:regionPloidAntag}). Thus, 
\vspace{0.5cm}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Both neo-W haplotypes can spread, in the absence of haploid selection, when there is overdominance
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
%\centerline{\includegraphics[width=1.25\linewidth]{Region_plot_combined_Mike}}
\centerline{\includegraphics[width=1.4\linewidth, right]{../Plots/Region_plot_combined.eps}}
\caption{
{\bf When the ancestral XY locus is tightly linked to a locus under selection ($r=0$), one or both neo-W haplotypes can spread (no haploid selection).}
We vary the fitness of male homozygotes relative to heterozygotes ($w_{Aa}^\circ=1$) and only consider stable equilibria at which both $\mathbf{A}$ locus alleles are maintained and the $a$ allele is initially fixed on the Y (non-hatched region). 
Here, selection in females can favour the $A$ allele (panel A, $w_{aa}^\female=0.85$, $w_{AA}^\female=1.05$), favour the $a$ allele (panel B, $w_{aa}^\female=1.05$, $w_{AA}^\female=0.85$), or be overdominant (panel C, $w_{aa}^\female=w_{AA}^\female=0.6$). 
If either haplotypic growth rate ($\Lambda_{W'A}^{(XY)}$ or $\Lambda_{W'a}^{(XY)}$) is greater than one, then a rare neo-W allele can spread for, at least, some values of $R>r$ (grey regions). 
The parameter values marked with an asterisk correspond to the fitnesses used in Fig \ref{fig:SexAntagTighter}C. 
%Where both haplotypic growth rates are greater than one, a neo-W will spread when rare, regardless of linkage with the selected locus (for any $R$). 
\nameref{fig:positionOverdominance} shows the dynamics arising with the parameters marked with a dagger. 
%Here, there is no haploid selection $t^\circ = \alpha^\circ_\Delta = 0$.
}
\label{fig:regionplots}
\end{figure}

In some cases, both neo-W-$A$ and neo-W-$a$ haplotypes can spread.
For example, when $AA$ individuals have low fitness in females yet the $A$ is polymorphic or fixed on the X background due to overdominance in males (Fig \ref{fig:regionplots}B and \ref{fig:regionplots}C), both neo-W-$A$ and neo-W-$a$ haplotypes produce fewer unfit $AA$ females.
This is true for the neo-W-$A$ haplotype because it can pair with a Y-$a$ haplotype and still be female. 
Whenever both haplotypic growth rates are greater than one, invasion by a neo-W is expected regardless of its linkage with the selected locus (i.e., for any $R$), see \nameref{fig:positionOverdominance} and \nameref{fig:temporalOverdominance} for examples. 
As a consequence, evolution can favor a new sex determination system on a different chromosome ($R = 1/2$), despite the fact that this unlinks the sex-determining locus from the selected locus.

@@

\noindent
With haploid selection, Conclusions 1 \& 2 continue to apply (\nameref{app:inv_cond}). 
The parameters for which neo-W-$A$ and neo-W-$a$ haplotypes spread under various forms of haploid selection are plotted in \nameref{fig:regionMaleDrive}, \nameref{fig:regionMaleGS}, \nameref{fig:regionFemaleDrive}, \nameref{fig:regionFemaleGS}.
In particular, we note that adding haploid selection allows shifts in sex determination to a different chromosome ($R=1/2$) even when selection is sexually antagonistic with directional selection in each diploid sex, e.g., \nameref{fig:SexAntagTighterMaleDrive}.  
Furthermore, haploid selection allows variation to be maintained by ploidally-antagonistic selection, under which trans-GSD transitions may also be favoured, \nameref{fig:regionPloidAntag}.
Some cases of XY $\rightarrow$ ZW transitions where $r=0$, $R=1/2$, and selection is ploidally-antagonistic (meiotic drive in males opposed by diploid selection) were studied by Kozielska et al. \cite{Kozielska:2010vm}, who found that sex-ratio biases are reduced during these transitions. 
However, such transitions are not always driven by selection to reduce sex-ratio bias. 
For example, with XY sex determination and haploid selection in females, sex ratios are not ancestrally biased yet a neo-W can invade (\nameref{fig:regionPloidAntag}). 
We further discuss how the spread of neo-sex-determining alleles is influenced by associations with haploid selected loci in the next section.

@@

@@

%\setcounter{equation}{0}
%\renewcommand{\theequation}{S1.\arabic{equation}}

>>>>

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

@@

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage[outdir=./]{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
%\fancyheadoffset[L]{2.25in}
%\fancyfootoffset[L]{2.25in}
%\lfoot{\sf PLOS}
%% Include all macros below

@@

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Haploid selection, sex ratio bias, and transitions between sex-determining systems} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Michael Francis Scott\textsuperscript{1\Yinyang*},
Matthew Miles Osmond\textsuperscript{2\Yinyang},
Sarah Perin Otto\textsuperscript{2}
\\
\bigskip
\textbf{1} UCL Genetics Institute, Department of Genetics, Evolution and Environment, University College London, Gower Street, London WC1E 6BT 
\\
\textbf{2} Department of Zoology, University of British Columbia, \#4200 - 6270 University Boulevard, Vancouver, BC, Canada V6T 1Z4
\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
%currently 312 words
\section*{Abstract}
Sex determination is remarkably dynamic; many taxa display shifts in the location of sex-determining loci or the evolution of entirely new sex-determining systems. 
Predominant theories for why we observe such transitions generally conclude that novel sex-determining systems are favoured by selection if they equalise the sex ratio or increase linkage with a locus that experiences different selection in males vs.\ females. 
We use population genetic models to extend these theories in two ways: 
(1) We consider the dynamics of loci very tightly linked to the ancestral sex-determining loci, e.g., within the non-recombining region of the ancestral sex chromosomes. 
Variation at such loci can favour the spread of new sex-determining systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or even unlinked) to the locus under selection. %, which is \textcolor{red}{not expected from previous theory}. 
(2) We consider selection upon haploid genotypes either during gametic competition (e.g., pollen competition) or meiosis (i.e., non-Mendelian segregation), which can cause the zygotic sex ratio to become biased. 
Haploid selection can drive transitions between sex-determining systems without requiring selection to act differently in diploid males vs.\ females. 
With haploid selection, we find that transitions between male and female heterogamety can evolve where linkage with the sex-determining locus is either strengthened or weakened. 
%Furthermore, we find that the selective strengths exerted by the sex-ratio and associations with selected alleles are often equivalent. 
Furthermore, we find that sex-ratio biases may increase or decrease with the spread of new sex chromosomes, which implies that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio.
In fact, under many conditions, we find that transitions in sex determination are favoured equally strongly in cases where the sex ratio bias increases or decreases.
%Furthermore, we find that the selective forces to equalize the sex ratio are equally important to selective forces that generate skewed sex ratios when accounting for the spread of new sex chromosomes.
%This allows sex-ratio biases to increase or decrease with the spread of new sex chromosomes and implies that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio. 
Overall, our models predict that sex determination systems should be highly dynamic, particularly when haploid selection is present, consistent with the evolutionary lability of this trait in many taxa.
%Overall, our models reveal that transitions between sex-determining systems, particularly transitions where the heterogametic sex changes, can be driven by loci in previously unpredicted genomic locations that experience selection during diploid and/or haploid phases.
%Our results predict conditions under which sex-determining systems are likely to be labile and draw connections with sex-ratio evolution.

@@

In each generation, we census the genotype frequencies in male and female gametes before gametic competition. 
%Below, it is convenient to follow the ancestral equilibrium frequency of the $A$ allele among female gametes (eggs), $\hat{p}^\female_X$, and among X-bearing, $\hat{p}^\male_X$, and among Y-bearing, $\hat{p}^\male_Y$, male gametes (sperm/pollen). 
%We also track the ancestral fraction of male gametes that are Y-bearing, $\hat{q}_{Y}$, which may deviate from $1/2$ due to meiotic drive in males. 
%
After gametic competition, conjugation between male and female gametes occurs at random.
The resulting zygotes develop as males or females, depending on their genotypes at the $\mathbf{X}$ and $\mathbf{M}$ loci. %(and the $\mathbf{M}$ genotype of their mother in the case of maternal control).
Diploid males and females then experience viability and/or individual-based fertility selection, with relative fitnesses $w_{AA}^{\circ}$, $w_{Aa}^{\circ}$, and $w_{aa}^{\circ}$. %where $g$ is the diploid genotype at the $\mathbf{A}$ locus ($g \in \{AA, Aa, aa\}$).  
We do not consider fertility selection that depends on the mating partner, e.g., sexual selection with variation in choosiness. 
The next generation of gametes is produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination (i.e., an odd number of cross-overs) occurs between loci $\mathbf{X}$ and $\mathbf{A}$ with probability $r$, between loci $\mathbf{A}$ and $\mathbf{M}$ with probability $R$, and between loci $\mathbf{X}$ and $\mathbf{M}$ with probability $\rho$.
Any linear order of the loci can be modelled with appropriate choices of $r$, $R$, and $\rho$ (see Fig \ref{fig:model_outline}A and Table \ref{tab:chisubstitutions}). 
Our model is entirely deterministic and hence ignores chance fluctuations in allele frequencies due to genetic drift.
%During the , the $\mathbf{A}$ locus can experience sex-specific gametic competition, diploid selection, and/or meiotic drive.

@@

In the ancestral population, it is convenient to follow the frequency of the $A$ allele among female gametes (eggs), $p^\female_X$, and among X-bearing, $p^\male_X$, or among Y-bearing, $p^\male_Y$, male gametes (sperm/pollen). 
We also track the fraction of male gametes that are Y-bearing, $q$, which may deviate from $1/2$ due to meiotic drive in males. 
We consider only equilibrium frequencies of alleles, $\hat{p}^\circ_i$, and Y-bearing male gametes, $\hat{q}_{Y}$, when determining the invasion of new sex-determining factors.
We use $\zeta$ to measure the sex ratio (fraction male) among zygotes, which is determined by the allele frequencies and haploid selection coefficients (Table \ref{tab:meanfitnesses}).

@@

\begin{table}[!ht]
%\centering
\smallskip
%\caption{Parameters determining invasion (equation \ref{eq:charpoly_neoY}) for neo-Y or neo-W alleles}
\caption{ {\bf Parameters determining invasion of mutant neo-Y and neo-W alleles into an ancestrally XY system}}
\begin{tabular}{l}
\hline\hline
   \noalign{\vskip 0.5ex}
   $m$ is a neo-Y ($k=0$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\Lambda_{Y'A}^{(XY)} = {\left( 2 \zeta \right)}^{-1} \left[\hat{p}_X^\female w_{A}^{\female} w_{A}^{\male} w_{AA}^{\male} + (1-\hat{p}_X^\female) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} (1+\alpha_{\Delta}^{\male}) \right]/ \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\male}_{D} \right) $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\Lambda_{Y'a}^{(XY)} = {\left( 2 \zeta \right)}^{-1} \left[(1-\hat{p}_X^\female) w_{a}^{\female} w_{a}^{\male} w_{aa}^{\male} + \hat{p}_X^\female w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male}(1 - \alpha_{\Delta}^{\male}) \right]/ \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\male}_{D} \right) $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{Y'A}^{(XY)} = R {\left( 2 \zeta \right)}^{-1} \left[ (1-\hat{p}_X^\female) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} (1+\alpha_{\Delta}^{\male}) \right]/  \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\male}_{D} \right)   $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{Y'a}^{(XY)} = R {\left( 2 \zeta \right)}^{-1} \left[   \hat{p}_X^\female w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male} (1 - \alpha_\Delta^{\male}) \right]/ \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\male}_{D} \right)  $\\ [1ex] \hline 
  \noalign{\vskip 0.5ex}
  $m$ is a neo-W ($k=1$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\Lambda_{W'A}^{(XY)} = {\left[ 2 (1 - \zeta) \right]}^{-1} \left[ \bar{p}^{\male} w_{A}^{\male} w_{A}^{\female} w_{AA}^{\female}+(1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female}(1+\alpha_{\Delta}^{\female})\right]/ \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\female}_{D} \right) $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\Lambda_{W'a}^{(XY)} = {\left[ 2 (1 - \zeta) \right]}^{-1} \left[ (1-\bar{p}^{\male}) w_{a}^{\male} w_{a}^{\female} w_{aa}^{\female}+\bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female}(1-\alpha_{\Delta}^{\female})\right] / \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\female}_{D} \right) $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{W'A}^{(XY)} = R {\left[ 2 (1 - \zeta) \right]}^{-1} \left[ (1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female} (1+\alpha_{\Delta}^{\female}) \right] / \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\female}_{D} \right)  $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{W'a}^{(XY)} = R {\left[ 2 (1 - \zeta) \right]}^{-1} \left[ \bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female} (1-\alpha_{\Delta}^{\female}) \right] / \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\female}_{D} \right) $ \\ [1ex]
  \hline \hline 
   \end{tabular}
   \begin{flushleft} 
$\bar{p}^{\male}=(1-\hat{q}_{Y})\hat{p}_X^\male + \hat{q}_{Y}\hat{p}_Y^\male$ is the average frequency of the $A$ allele among X- and Y-bearing male gametes.
$\hat{q}_{Y}$ is the frequency of Y-bearing male gametes. 
%      \item $R$ is the probability of recombination between loci $\mathbf{A}$ and $\mathbf{M}$.
$\zeta$ is the zygotic sex ratio (fraction male).
$\bar{w}^{\circ}_{D}$ is the mean fitness of diploids of sex $\circ$ $\in\{\female,\male\}$.
$\bar{w}_H^\circ$ is the mean fitness of haploids from sex $\circ$, see Table \ref{tab:meanfitnesses}.
$R$ is the rate of recombination between the neo-sex-determiner and the selected locus.
Selection terms ($w_i^\circ$, $\alpha_\Delta^\circ$) are described in Table \ref{tab:fitnesstable}. 
\end{flushleft}
  \label{tab:haplotype_growth}
\end{table}

@@

Neo-W alleles, on the other hand, can invade an ancestral XY system (the full invasion conditions are given in \nameref{app:inv_cond}; equations \ref{Binvasion} and \ref{Ainvasion}).
Invasion occurs when neo-W females can have higher fitness than the XX females in the ancestral population.
Neo-W invasion is possible under all forms of selection that can maintain a polymorphism (sexually-antagonistic selection, overdominance, ploidally-antagonistic selection, or some combination, e.g., Fig \ref{fig:temporalOverdominance}, Fig \ref{fig:SexAntagTighterMaleDrive}, and Fig \ref{fig:regionPloidAntag}). Thus, 
\vspace{0.5cm}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Both neo-W haplotypes can spread, in the absence of haploid selection, when there is overdominance
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
%\centerline{\includegraphics[width=1.25\linewidth]{Region_plot_combined_Mike}}
\centerline{\includegraphics[width=1.2\linewidth]{../Plots/Region_plot_combined.eps}}
\caption{
{\bf When the ancestral XY locus is tightly linked to a locus under selection ($r=0$), one or both neo-W haplotypes can spread (no haploid selection).}
We vary the fitness of male homozygotes relative to heterozygotes ($w_{Aa}^\circ=1$) and only consider stable equilibria at which both $\mathbf{A}$ locus alleles are maintained and the $a$ allele is initially fixed on the Y (non-hatched region). 
Here, selection in females can favour the $A$ allele (panel A, $w_{aa}^\female=0.85$, $w_{AA}^\female=1.05$), favour the $a$ allele (panel B, $w_{aa}^\female=1.05$, $w_{AA}^\female=0.85$), or be overdominant (panel C, $w_{aa}^\female=w_{AA}^\female=0.6$). 
If either haplotypic growth rate ($\Lambda_{W'A}^{(XY)}$ or $\Lambda_{W'a}^{(XY)}$) is greater than one, then a rare neo-W allele can spread for, at least, some values of $R>r$ (grey regions). 
The parameter values marked with an asterisk correspond to the fitnesses used in Fig \ref{fig:SexAntagTighter}C. 
%Where both haplotypic growth rates are greater than one, a neo-W will spread when rare, regardless of linkage with the selected locus (for any $R$). 
Fig \ref{fig:positionOverdominance} shows the dynamics arising with the parameters marked with a dagger. 
%Here, there is no haploid selection $t^\circ = \alpha^\circ_\Delta = 0$.
}
\label{fig:regionplots}
\end{figure}

In some cases, both neo-W-$A$ and neo-W-$a$ haplotypes can spread.
For example, when $AA$ individuals have low fitness in females yet the $A$ is polymorphic or fixed on the X background due to overdominance in males (Fig \ref{fig:regionplots}B and \ref{fig:regionplots}C), both neo-W-$A$ and neo-W-$a$ haplotypes produce fewer unfit $AA$ females.
This is true for the neo-W-$A$ haplotype because it can pair with a Y-$a$ haplotype and still be female. 
Whenever both haplotypic growth rates are greater than one, invasion by a neo-W is expected regardless of its linkage with the selected locus (i.e., for any $R$), see Fig \ref{fig:positionOverdominance} and Fig \ref{fig:temporalOverdominance} for examples. 
As a consequence, evolution can favor a new sex determination system on a different chromosome ($R = 1/2$), despite the fact that this unlinks the sex-determining locus from the selected locus.

@@

\noindent
With haploid selection, Conclusions 1 \& 2 continue to apply (\nameref{app:inv_cond}). 
The parameters for which neo-W-$A$ and neo-W-$a$ haplotypes spread under various forms of haploid selection are plotted in Fig \ref{fig:regionMaleDrive}, Fig \ref{fig:regionMaleGS}, Fig \ref{fig:regionFemaleDrive}, Fig \ref{fig:regionFemaleGS}.
In particular, we note that adding haploid selection allows shifts in sex determination to a different chromosome ($R=1/2$) even when selection is sexually antagonistic with directional selection in each diploid sex, e.g., Fig \ref{fig:SexAntagTighterMaleDrive}.  
Furthermore, haploid selection allows variation to be maintained by ploidally-antagonistic selection, under which trans-GSD transitions may also be favoured, Fig \ref{fig:regionPloidAntag}.
Some cases of XY $\rightarrow$ ZW transitions where $r=0$, $R=1/2$, and selection is ploidally-antagonistic (meiotic drive in males opposed by diploid selection) were studied by Kozielska et al. \cite{Kozielska:2010vm}, who found that sex-ratio biases are reduced during these transitions. 
However, such transitions are not always driven by selection to reduce sex-ratio bias. 
For example, with XY sex determination and haploid selection in females, sex ratios are not ancestrally biased yet a neo-W can invade (Fig \ref{fig:regionPloidAntag}). 
We further discuss how the spread of neo-sex-determining alleles is influenced by associations with haploid selected loci in the next section.

@@

\section*{Acknowledgments}
We thank Georgy Sandler and Stephen Wright for sharing their results with us, and we thank Bret Payseur and three anonymous reviewers for helpful comments on this manuscript.

% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% or
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here. See http://journals.plos.org/plosone/s/latex for 
% step-by-step instructions.
% 
\bibliography{sex_chromosomes.bib}

@@

\setcounter{equation}{0}
\renewcommand{\theequation}{S2.\arabic{equation}}
